---
title: "More information"
author: "Melanie N, Ratie, Joan K"
date: "Last updated on `r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: kable
---

<!--
Comments in HTML are like this! 
-->

```{r, echo = FALSE, error = FALSE, message = FALSE, warning = FALSE}
# Load the required libraries
library(readr)
library(tidyverse)
library(openintro)
library(maps)
library(scales)

# Import datasets from Buzzfeed Fake News survey
headline_responses <- read_csv("headline-responses.csv")
raw_data <- read_csv("raw-data.csv")
```

<!--
CREATE BASE MAP THAT SPLITS THE US INTO REGIONS FOR VISUALIZATION 1
Adapted from: https://gis.stackexchange.com/questions/141181/how-to-create-a-us-map-in-r-with-separation-between-states-and-clear-labels
-->

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Load base map from maps package
us.map <-  map_data('state')

# Add zones: Northeast, Southwest, Southeast, East, Midwest
us.map$zone[us.map$region %in% 
              c("maine", "vermont", "new hampshire", "massachusetts",
                "connecticut", "rhode island", "new york", "pennsylvania",
                "new jersey", "delaware", "district of columbia",
                "maryland")] <- "zone 1: Northeast"

us.map$zone[us.map$region %in% 
              c("south dakota", "north dakota","nebraska", "kansas",
                "minnesota", "iowa", "missouri", "wisconsin",
                "illinois", "indiana", "michigan", "ohio")] <- "zone 2: Midwest"

us.map$zone[us.map$region %in% 
              c("oklahoma", "new mexico", "texas", "arizona")] <- "zone 3: Southwest"

us.map$zone[us.map$region %in% 
              c("montana", "idaho", "wyoming", "utah",
                "colorado", "washington", "oregon", "nevada",
                "california")] <- "zone 4: West"

us.map$zone[us.map$region %in% 
              c("kentucky", "mississippi","tennessee", 
                "arkansas", "louisiana", "alabama", 
                "west virginia", "virginia", "north carolina",
                "south carolina", "georgia", "florida")] <- "zone 5: Southeast"

# Subset the dataframe by  zones and move lat/lon accordingly
us.map$lat.transp[us.map$zone == "zone 1: Northeast"] <- 
  us.map$lat[us.map$zone == "zone 1: Northeast"] 
us.map$long.transp[us.map$zone == "zone 1: Northeast"] <- 
  us.map$long[us.map$zone == "zone 1: Northeast"] + 5

us.map$lat.transp[us.map$zone == "zone 2: Midwest"] <- 
  us.map$lat[us.map$zone == "zone 2: Midwest"]
us.map$long.transp[us.map$zone == "zone 2: Midwest"] <- 
  us.map$long[us.map$zone == "zone 2: Midwest"]

us.map$lat.transp[us.map$zone == "zone 3: Southwest"] <- 
  us.map$lat[us.map$zone == "zone 3: Southwest"] - 3
us.map$long.transp[us.map$zone == "zone 3: Southwest"] <- 
  us.map$long[us.map$zone == "zone 3: Southwest"] 

us.map$lat.transp[us.map$zone == "zone 4: West"] <- 
  us.map$lat[us.map$zone == "zone 4: West"]
us.map$long.transp[us.map$zone == "zone 4: West"] <-
  us.map$long[us.map$zone == "zone 4: West"] - 5

us.map$lat.transp[us.map$zone == "zone 5: Southeast"] <-
  us.map$lat[us.map$zone == "zone 5: Southeast"] - 3
us.map$long.transp[us.map$zone == "zone 5: Southeast"] <- 
  us.map$long[us.map$zone == "zone 5: Southeast"] + 5

# Add state abbreviations to appear as labels on map
states <- 
  aggregate(cbind(long.transp, lat.transp) ~ region, data=us.map,
            FUN=function(x)mean(range(x))) %>%
  mutate(labels = state2abbr(region))
```

<!--
DATA WRANGLING OF BUZZFEED SURVEY RESULTS
-->

```{r, error = FALSE, message = FALSE, warning = FALSE}
# Select the needed columns for merging 
raw_data <- raw_data %>%
  select(ID, HCAL_REGION1_Label_US, resp_gender, resp_age, DWD4, DWD5, DWD6, DWD1)

# Inner join the 2 datasets
joined_data <- headline_responses %>%
  inner_join(raw_data, by = "ID")

# Data wrangling for the joined datset
joined_data <- joined_data %>%
  # Select only instances where respondents recall a news headline
  filter(recalled == "yes") %>%
  # rename the variables
  rename( US_states = HCAL_REGION1_Label_US,
          party_affiliation = DWD4,
          if_voted = DWD5,
          political_party = DWD1) %>%
  # create new variables to reflect age group, political party, president choice
  mutate(age_group = case_when(resp_age %in% c(94:108) ~ "GREATEST GENERATION",
                               resp_age %in% c(73:93) ~ "SILENT GENERATION",
                               resp_age %in% c(72:54) ~ "BABY BOOMERS",
                               resp_age %in% c(39:53) ~ "GEN X",
                               TRUE ~ "MILLENIALS"),
         age_group = fct_relevel(age_group, 
                                 "GREATEST GENERATION",
                                 "SILENT GENERATION",
                                 "BABY BOOMERS",
                                 "GEN X",
                                 "MILLENIALS"),
         political_party = case_when(political_party == "1" ~ "Democrat",
                                     political_party == "2" ~ "Republican",
                                     political_party == "3" ~ "Independent",
                                     political_party == "4" ~ "Other"),
         political_party = fct_relevel(political_party,
                                       "Republican",
                                       "Democrat",
                                       "Independent",
                                       "Other"),
         DWD6 = replace_na(DWD6, 0),
         presidential_candidate = case_when(DWD6 == 1 ~ "Hillary Clinton",
                                            DWD6 == 2 ~ "Donald Trump",
                                            DWD6 == 3 ~ "Gary Johnson",
                                            DWD6 == 4 ~ "Jill Stein",
                                            DWD6 == 5 ~ "Other",
                                            DWD6 == 0 ~ "Did not vote"),
         presidential_candidate = fct_relevel(presidential_candidate,
                                              "Donald Trump",
                                              "Hillary Clinton",
                                              "Gary Johnson",
                                              "Jill Stein",
                                              "Other",
                                              "Did not vote"),
         # Create variable wrong_or_not that determines if the respondent 
         # judge a headline correctly. 1 for wrong, 0 for correct
         wrong_or_not = ifelse(accuracy_bool == is_fake, 1, 0),
         # Create variable abbr for state abbreviation
         abbr = state2abbr(US_states),
         # Format US_states to lower case
         US_states = tolower(US_states))

# Create new dataframe with percentage headlines judged wrongly per respondent
avg_wrong_responses <- joined_data %>%
  # Calculate the percentage of wrong responses
  group_by(ID, Weightvar, US_states, abbr, resp_gender, 
           age_group, presidential_candidate, political_party) %>%
  summarize(wrong_count = sum(wrong_or_not),
         total = n(),
         percentage_wrong = wrong_count/total)

# Data wrangling of for Visualization 1: choropleth map
map_data <- avg_wrong_responses %>%
  group_by(US_states, abbr) %>%
  summarize(state_average = weighted.mean(percentage_wrong, Weightvar),
            state_average_discrete = cut(state_average,
                                         breaks = c(0.1, 0.2, 0.3, 0.4, 0.5, 1),
                                         labels = c("10% - 20%", "20% - 30%",
                                                    "30% - 40%", "40% - 50%", "Over 50%"))) %>%
  inner_join(us.map, by = c("US_states" = "region"))

# Data wrangling for Visualization 2: Plot of political bias by age group & political party 
# Group by age groups and political party
wrong_per_age <- avg_wrong_responses %>%
  # Calculate the percentage of wrong responses
  group_by(age_group, political_party) %>%
  mutate(mean_wrong = weighted.mean(percentage_wrong, Weightvar),
         mean_wrong = round(mean_wrong, digits = 2))
#Calculate the mean percentage wrong for each generation
perc_wrong_per_age <- wrong_per_age %>%
  group_by(age_group) %>%
  summarize(gen_wrong = mean(percentage_wrong)) %>%
  mutate(gen_wrong = round(gen_wrong, digits = 2))

# Data wrangling for Visualization 3:
# Select for headline C and K only
presidential_data <- joined_data %>%
  filter (headline == c("C", "K")) %>%
  group_by(headline, wrong_or_not, presidential_candidate) %>%
  count(presidential_candidate)

```

<!--
MAIN BODY STARTS HERE
-->

##Bias By Geographic Region

The following choropleth map gives an overview of how much an average person from each state misjudge news headlines. States are split up by their respective regions (in clockwise order from the top left: West, Midwest, Northeast, Southeast, and Southwest). Interestingly, Utah is the state where people seem to have the best perception of news accuracy. In contrast, the five states with the highest percentage of misjudged news are Wyoming, Kansas, New Mexico, Vermont and Alabama. These are a mixture of red and blue state, and each belonging to one of the five different regions.  
Clearly, no matter which region of the US we are looking at, people are prone to misjudge headlines they see. Thus, we decided to look deeper into the survey data and examine the individual characteristics of the respondents in an attempt to determine what type of person is more vulnerable to fake news.  

```{r, error = FALSE, message = FALSE, warning = FALSE, fig.width = 10}
# Plot choropleth map
ggplot(map_data,  
       aes(x = long.transp, y = lat.transp), colour = "white") + 
  geom_polygon(aes(group = group, fill = state_average_discrete)) +
  geom_text(data = states, aes(long.transp, lat.transp, label = labels), 
            size=2) +
  labs(title = 'What percentage of news headlines does an average person misjudge in each state?',
       fill = 'Percentage') +
  scale_fill_brewer(palette = "YlGnBu") +
  theme_bw() +
  theme(
    text = element_text(size = 10, family = "Roboto"),
    legend.text=element_text(size = 8, family = "Roboto"),
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    panel.grid = element_blank(), 
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank())
```


##Bias By Political Party Affiliation Across Age Groups

As the findings from this [scatterplot](/index.html)  suggested, the more strongly republican a state is, the more susceptible they were to believing fake news and the opposite applied to Democratic states. We sought to find out if there was any correlation between the percentage of headlines respondents got wrong (i.e. either fake news believed to be true or real news believed to be false) and the political party they associated with (taking into account non-Democratic and non-Republican parties) across different age groups.   
 
Our dataset contained information for:
1619 Democrats   
1465 Republicans   
952 respondents affiliated with Other parties and   
99 individuals affiliated with Independent parties.   

We selected data for the respondents who recalled seeing certain headlines and split the respondents into 5 age groups.   

The greatest generation had the smallest population of 6 people and millenials was the biggest age group with 1899 total respondents. The Silent generation had 180 respondents, the Baby Boomer generation had 1134 people and lastly Gen X had 916 respondents.

A comparison of total bias across the age groups confirmed that Republicans and respondents affiliated with “Other” political Parties generally had the highest percentage of wrong responses across the age groups, followed by Democrats and lastly respondents affiliated with Independent Parties. 

```{r, error = FALSE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 8}
# Create a plot that shows bias by Political party affiliation across age groups
ggplot() +
  geom_col(data = wrong_per_age, position="dodge", 
           aes(x = age_group, y = mean_wrong, fill = political_party)) +
  geom_boxplot(data = perc_wrong_per_age, 
               aes(x = age_group, y = gen_wrong)) +
  geom_text(data = perc_wrong_per_age, 
            aes(label = paste(gen_wrong*100, '%'),
                x = age_group, y = gen_wrong + 0.03)) +
  labs(x = "Age group", y = "Percentage of Wrong Responses", 
       title = "Bias by Political party affiliatiton across age groups",
       fill = "Respondent's Political Party") +
  scale_fill_manual(values = c("Red","Blue","Purple","Gray")) +
  scale_y_continuous(labels = percent, breaks=seq(0, 1, .1)) +
  # change the title and axes size
  theme_bw() +
  theme(text = element_text(size = 10, family = "Roboto"),
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
        legend.text=element_text(size = 8, family = "Roboto"),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12))
```

The reason why Republicans had such high bias is probably because most of the headlines involving information on Donald Trump were mostly positive and those involving Hillary clinton were negative, hence Trump supporters were most likely to believe most of the news to be true and Democrats were most likely to be dubious. However, individuals associated with Other and independent parties and were not in support of either Donald Trump of Hillary Clinton were then more likely to be swayed in the Republican direction if they had no strong beliefs on whether or not the news was accurate.

The values indicated on the plot show the average percentage wrong for each age group and we can see that bias was high for the older age groups, then became low for the middle lying age groups, and then went high again for the younger groups.

We believe that the main cause of this is social media usage across the different age groups. According to Statista, Millenials are the largest social media users with 88% of Millennials using some form of social media platform everyday.  Only 37% of individuals above 65 years use social media and this could mean that most of them are not well informed about current political affairs and so become more susceptible to believing nearly anything they hear. However, Millenials use too much social media and chances of them accessing fake news are highly likely.

We further look into the differences, if any exists, between people who believed in fake news vs people who believed in real news and whom they voted for in the 2016 presidential elections. 

##Fake News & The Presidential Election

"Does believing in fake news have any correlation with a person's candidate choice in the 2016 Presidential Election?" is a question that we are interested in answering in this section. 

In answering this question, we decide to look at the most recalled fake news and true news headline. The most recalled fake headline is *FBI Agent Suspected in Hillary Email Leaks Found Dead in Apparent Murder Suicide.* and the most recalled true headline is *Donald Trump on Refusing Presidential Salary: "I'm not taking it"* 

We have made a bar plot that shows, for each of the selected headlines, the relationship between the accuracy of people's perception of the news and the presidential candidate they voted for. Our plot is faceted based on whether the news is fake or not, and whether people's perception are correct or not, so that we can see if fake news made any impact on the presidential candidate that was elected.

```{r, error = FALSE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 8}
# create a plot that shows the distibutions of elected presidents
# Based on whether the news was fake or real
ggplot(presidential_data) +
  geom_col(aes(x = presidential_candidate, y = n, fill = presidential_candidate)) +
  facet_grid(headline ~ wrong_or_not,
            labeller = labeller(headline = c("C" = "Fake news", "K" = "Real news"),
                                wrong_or_not = c("0" = "Correct perception", "1" = "Wrong perception"))) +
  # red - Republic, blue - Democrats
  # goldenrod4 - Libertarian party, darkolivegreen4 - green party
  # gray - others
  scale_fill_manual(values = c("Red", "Blue", "goldenrod4", "darkolivegreen4", "gray", "pink")) +
  # change the facet labels to fake news and real news
  labs(title = "Distribution of votes among presidential candidates", x = "Presidential candidates", y = "Number of votes") +
  theme_bw() +
  # change the text size of the axes and the title
  theme(text = element_text(size = 10, family ="Roboto"),
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12))
```

From the bar plot above, we see that for the headline that was real, out of the people were able to identify the news as accurate, most of them voted for Donald Trump. We also see that Donald Trump and Hillary Clinton were the top two leading presidential candidates with Hillary Clinton having only 30 votes less than Donald Trump.

However, for the news headlines that were fake and people identified these headlines as accurate, Donald Trump ended up having the most votes compared to the rest of the presidential candidates. Even though Donald Trump and Hillary Clinton were the top two leading presidential candidates with more votes, there is a huge difference between the two candidates with Donald Trump having more than twice the votes that Hillary Clinton had.   
The Fake news headlines include:   
A is *Pope Francis Shocks the World, Endorses Donald Trump for president, Releases Statement.*   
B is *Donald Trump Sent His Own Plane To Transport 200 Stranded Marines.*   
C is *FBI Agent Suspected in Hillary Email Leaks Found Dead in Apparent Murder Suicide.*  
D is *Donald Trump Protestor Speaks Out: I was paid $3.500 to Protest Trump's Rally.*   
E is *FBI Director Comey Just Put a Trump Sign On His Front Lawn.*   
From the fake news headlines above, we see that these headlines are all in favor of Donald Trump with headline like C considered an anti-Hillary Clinton story ^[https://www.huffingtonpost.com/entry/facebook-fake-news-stories-zuckerberg_us_5829f34ee4b0c4b63b0da2ea] as the news puts Hillary Clinton at a bad position to be doubted as to why the FBI was found dead after being suspected to leak her emails.   
Thus, we can assume that these people who believed the fake news headlines to be accurate were then in favor of having Donald Trump as president. Our assumption is not completely accurate as this may or may not have been one of the reasons that led to Donald Trump winning the presidential elections. Moreover, our sample size does not reflect the whole of the US but only the few respondents who completed the survey. However, fake news may have contributed directly or indirectly in either raising doubts among American voters, or changing some people's views of either having Hillary Clinton as president or changing their decisions to not wanting to vote for any president.
